---
title: "Heat-related Hospitalisations - Condition Breakdown"
output: 
  pdf_document:
    # toc: true
    number_sections: true
    fig_caption: true
    keep_tex: false
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

library(tidyverse)
library(phsstyles)
library(cowplot)

source("hosp_condition_analysis.R")
```

The heat-related hospitalisation causes that we include in our DLNM analysis can be divided into two groups: physiological and behavioural. The table below shows which causes are assigned to which group.

| Behavioural Conditions | Physiological Conditions |
|------------------------|--------------------------|
| Dehydration            | Alzheimer's              |
| Drowning               | Cardiovascular           |
| Falls                  | Dementia                 |
| Injuries               | Hot weather              |
| Mental health          | Parkinson's              |
| Natural forces         | Renal                    |
| Road incidents         | Respiratory              |
| Suicide & self-harm    |                          |
| Violence               |                          |

The following analysis looks at hospital admissions in Scotland in the years 2005-2024, only including the months June-September for each.

\newpage

```{r}
## Leaving 'count' plot here for reference/future use
# df %>%
#   group_by(year) %>%
#   summarise(Behavioural = sum(all_behavioural),
#             Physiological = sum(all_physiological),
#             Mixed = sum(mixed_conditions)) %>% 
#   pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
#   mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
#   ggplot() +
#   geom_line(aes(x = year, y = Count, group = `Condition breakdown`, colour = `Condition breakdown`),
#             linewidth = 1.2) +
#   scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
#   geom_point(aes(x = year, y = Count, group = `Condition breakdown`), colour = "black") +
#   theme_minimal() +
#   theme(
#     axis.line = element_line(color = "black")
#   )+
#   labs(
#     x = "Year"
#   )

## Proportion plot
df %>%
  group_by(year) %>%
  summarise(Hospitalisations = n(), # for reference
            Behavioural = sum(all_behavioural)/n()*100,
            Physiological = sum(all_physiological)/n()*100,
            Mixed = sum(mixed_conditions)/n()*100) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Proportion (%)") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_line(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`, colour = `Condition breakdown`),
            linewidth = 1.2) +
  scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
  geom_point(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`), colour = "black") +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  labs(
    x = "Year"
  )
```

In the summer months, hospitalisations due to exclusively physiological causes are consistently the most frequent.

\newpage

```{r}
## Leaving 'count' plot here for reference/future use
# df %>%
#   group_by(year, age_group) %>%
#   summarise(Behavioural = sum(all_behavioural),
#             Physiological = sum(all_physiological),
#             Mixed = sum(mixed_conditions)) %>% 
#   pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
#   mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
#   ggplot() +
#   geom_line(aes(x = year, y = Count, group = `Condition breakdown`, colour = `Condition breakdown`),
#             linewidth = 1.2) +
#   scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
#   geom_point(aes(x = year, y = Count, group = `Condition breakdown`), colour = "black",size = 0.5) +
#   theme_minimal() + 
#     theme(
#     axis.line = element_line(color = "black")
#   )+
#   facet_wrap(vars(age_group),
#              labeller = as_labeller(c(`under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))) +
#   labs(
#     x = "Year",
#     y = "Count\n"
#   )


## Proportion plot
df %>%
  group_by(year, age_group) %>%
  summarise(Hospitalisations = n(), # for reference
            Behavioural = sum(all_behavioural)/n()*100,
            Physiological = sum(all_physiological)/n()*100,
            Mixed = sum(mixed_conditions)/n()*100) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Proportion (%)") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_line(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`, colour = `Condition breakdown`),
            linewidth = 1.2) +
  scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
  geom_point(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`), colour = "black",size = 0.5) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  facet_wrap(vars(age_group),
             labeller = as_labeller(c(`under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))) +
  labs(
    x = "Year",
  )
```

However, when we split the data by age group, we see that younger people are hospitalised due to behavioural causes much more frequently than older people, while hospitalisations of older people are heavily skewed towards physiological causes.

\newpage

```{r}
## Leaving 'count' plot here for reference/future use
# df %>%
#   group_by(year, age_group, simd) %>%
#   summarise(Behavioural = sum(all_behavioural, na.rm = TRUE),
#             Physiological = sum(all_physiological, na.rm = TRUE),
#             Mixed = sum(mixed_conditions, na.rm = TRUE)) %>% 
#   pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>%
#   mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
#   filter(!is.na(simd)) %>% 
#   ggplot() +
#   geom_line(aes(x = year, y = Count, group = `Condition breakdown`, colour = `Condition breakdown`),
#             size = 0.05, linewidth = 1.2) +
#   scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
#   geom_point(aes(x = year, y = Count, group = `Condition breakdown`), colour = "black",size = 0.25) +
#   theme_minimal() + 
#   facet_grid(age_group ~ simd,
#              labeller = as_labeller(c(`1` = 'SIMD 1', `2` ='SIMD 2', `3` = 'SIMD 3', `4` ='SIMD 4', `5` = 'SIMD 5', `under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))) +
#   labs(
#     x = "Year"
#   ) +
#   theme(axis.text.x = element_text(angle = 45, vjust = 0.5),
#         axis.line = element_line(color = "black"))

## Proportion plot
df %>%
  group_by(year, age_group, simd) %>%
  summarise(Hospitalisations = n(), # for reference
            Behavioural = sum(all_behavioural)/n()*100,
            Physiological = sum(all_physiological)/n()*100,
            Mixed = sum(mixed_conditions)/n()*100) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Proportion (%)") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
  filter(!is.na(simd)) %>% 
  ggplot() +
  geom_line(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`, colour = `Condition breakdown`),
            size = 0.05, linewidth = 1.2) +
  scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
  geom_point(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`), colour = "black",size = 0.25) +
  theme_minimal() + 
  facet_grid(age_group ~ simd,
             labeller = as_labeller(c(`1` = 'SIMD 1', `2` ='SIMD 2', `3` = 'SIMD 3', `4` ='SIMD 4', `5` = 'SIMD 5', `under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))) +
  labs(
    x = "Year"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5),
        axis.line = element_line(color = "black"))
```

When we add deprivation index into the equation, we can see that this age pattern remains, but is strongest in the most deprived areas.

\newpage

```{r}
## Leaving 'count' plot here for reference/future use
# df %>%
#   group_by(year, age_group, sex) %>%
#   summarise(Behavioural = sum(all_behavioural, na.rm = TRUE),
#             Physiological = sum(all_physiological, na.rm = TRUE),
#             Mixed = sum(mixed_conditions, na.rm = TRUE)) %>% 
#   pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
#   mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
#   ggplot() +
#   geom_line(aes(x = year, y = Count, group = `Condition breakdown`, colour = `Condition breakdown`),
#             linewidth = 1.2) +
#   scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
#   geom_point(aes(x = year, y = Count, group = `Condition breakdown`), colour = "black",
#              size = 0.3) +
#   theme_minimal() +
#   theme(
#     axis.line = element_line(color = "black")
#   )+
#   facet_grid(age_group ~ sex,
#              labeller = as_labeller(c(female='Female', male='Male',
#                                       `under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))
#   )  +
#   labs(
#     x = "Year"
#   ) +
#   theme(axis.text.x = element_text(angle = 45, vjust = 0.5))

## Proportion plot
df %>%
  group_by(year, age_group, sex) %>%
  summarise(Hospitalisations = n(), # for reference
            Behavioural = sum(all_behavioural)/n()*100,
            Physiological = sum(all_physiological)/n()*100,
            Mixed = sum(mixed_conditions)/n()*100) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Proportion (%)") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels=c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_line(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`, colour = `Condition breakdown`),
            linewidth = 1.2) +
  scale_colour_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta")))+
  geom_point(aes(x = year, y = `Proportion (%)`, group = `Condition breakdown`), colour = "black",
             size = 0.3) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  facet_grid(age_group ~ sex,
             labeller = as_labeller(c(female='Female', male='Male',
                                      `under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" ))
  )  +
  labs(
    x = "Year"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

There isn't much of a difference between sexes in admission causes in older people, but for younger people we can clearly see young men are involved in considerably more behavioural-related admissions.

\newpage

```{r}
df %>%
  group_by(temperature_group) %>%
  summarise(Behavioural = sum(all_behavioural, na.rm = TRUE),
            Physiological = sum(all_physiological, na.rm = TRUE),
            Mixed = sum(mixed_conditions, na.rm = TRUE)) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels = c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_col(aes(x = temperature_group, y = Count, fill = `Condition breakdown`), colour = "black") +
  scale_fill_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta"))) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  labs(
    x = "\nDaily max temperature group"
  )
```

When we split the data into bins based on daily maximum temperatures, we can see that most admissions fall within the 15-20 category due to these temperatures being the most common. If we then look at condition proportions, however, we see that these remain highly consistent across temperature categories.

```{r}
df %>%
  group_by(temperature_group) %>%
  summarise(Behavioural = sum(all_behavioural, na.rm = TRUE),
            Physiological = sum(all_physiological, na.rm = TRUE),
            Mixed = sum(mixed_conditions, na.rm = TRUE)) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels = c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_col(aes(x = temperature_group, y = Count, fill = `Condition breakdown`), colour = "black",
           position = "fill", show.legend = TRUE) +
  scale_fill_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta"))) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  labs(
    x = "\nDaily max temperature group",
    y = "Proportion\n"
  )

```

\newpage

```{r}
df %>% 
  group_by(date) %>% 
  summarise(temp = mean(max_temperature),
            total_behavioural = sum(all_behavioural),
            total_physiological = sum(all_physiological),
            total_mixed = sum(mixed_conditions)) %>% 
  pivot_longer(cols = contains("total"), names_to = "condition_category", values_to = "total") %>% 
  ggplot(aes(x = temp, y = total, colour = condition_category)) +
  geom_point(alpha = 0.3, show.legend = FALSE) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black")
  )+
  facet_wrap(vars(condition_category), nrow = 3,
             labeller = as_labeller(c(total_behavioural = 'Behavioural causes', total_physiological = 'Physiological causes', total_mixed = 'Mixed causes'))
  ) +
  labs(
    x = "\nMean daily temperature (°C)",
    y = "Count\n",
    colour = "Condition category"
  )
```

This finding is reinforced when we examine the relationship between daily temperatures and admission counts using scatterplots: no admission category shows a discernible trend. Note that each day's temperature is calculated as the mean of all maximum temperatures recorded in locations with hospital admissions.

\newpage

```{r}
df %>%
  group_by(temperature_group, age_group, simd) %>%
  summarise(Behavioural = sum(all_behavioural, na.rm = TRUE),
            Physiological = sum(all_physiological, na.rm = TRUE),
            Mixed = sum(mixed_conditions, na.rm = TRUE)) %>% 
  filter(!is.na(simd)) %>% 
  pivot_longer(cols = c(Behavioural, Physiological, Mixed), names_to = "Condition breakdown", values_to = "Count") %>% 
  mutate(`Condition breakdown` = factor(`Condition breakdown`, levels = c("Behavioural", "Physiological", "Mixed"))) %>% 
  ggplot() +
  geom_col(aes(x = temperature_group, y = Count, fill = `Condition breakdown`), colour = "black",
           position = "fill") +
  scale_fill_manual(values=c(Behavioural = phs_colours("phs-blue-50"), Physiological = phs_colours("phs-green"), Mixed = phs_colours("phs-magenta"))) +
  theme_minimal() +  theme(
    axis.line = element_line(color = "black")
  )+
  labs(
    x = "\nDaily max temperature group",
    y = "Proportion\n"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(age_group ~ simd,
             labeller = as_labeller(c(`1` = 'SIMD 1', `2` ='SIMD 2', `3` = 'SIMD 3', `4` ='SIMD 4', `5` = 'SIMD 5', `under 65 yrs` = "< 65yrs", `65 yrs plus` = "65yrs +" )))



```

Splitting the proportion results up by age group and deprivation level shows similar demographic patterns to the previous results, but not much change between temperature levels.
