```{r setup, include=FALSE}
library(here)
library(knitr)
# source(here("dlnm/dlnm_setup.R")) # have commented out while sourcing from main script
# For df:
# source(here("dlnm/dlnm_code.R")) # have commented out while sourcing from main script
# For knots:
#source(here("dlnm/dlnm_code_knots.R"))

knitr::opts_chunk$set(echo = TRUE)
```
---
title: "Heat-related `r event` (`r age_split_title`)"
author: "Bella Tortora Brayda / Sarah Reed / Al Morgan"
date: `r today_date`
output: html_document
---

## DLNM Model Evaluation

### Context
 
A Distributed Lag Non-Linear Model (DLNM) is a prominent method used in population health to determine the effect of an exposure on a given outcome, allowing for a delay between exposure to an event and the resulting impact. 

In our context, we are evaluating how `r exposure` impacts `r event`.

For us to be able to accurately calculate the relative risk of `r event` from `r exposure`, and to assess the attributable `r event` for a given timeframe, we need a model that is robust. 

`r indicator_context`

This R markdown compiles the outputs created in the dlnm_code, and is designed to facilitate evaluating which, if any, of the model variations, produces robust estimates for attributable `r event` to `r exposure`.

## Parameters and variables of interest

This model variation evaluates the following independent columns: 

* `r independent_col1`
* `r independent_col2`
* `r independent_col3`
* `r independent_col4`

The parameters are as follows:

* Event = `r event`
* Exposure = `r exposure`
* Exposure function (see dlnm::crossbasis) = `r varfun`
* Months included = `r month_choice`
* Degrees of freedom in exposure function (see dlnm::crossbasis) = `r vardegree` 
* Lag length in days = `r lag` 
* Degrees of freedom in lag function = `r lagdf`
* Degrees of freedom for seasonality = `r dfseas` 
* Percentile used for initially centring predictions = `r percentile` 

## Outputs of interest:

The main things to look for in the below output are:

1) Significance across the time periods (y1-y5)

This is indicated by small p-values (Pr(>|z|)) and for quick reference R signals these with stars. These indicate which variables significantly impact `r event` in our system.

2) Multivariate Cochran Q-test. 

This tells you the probability that the observed level of heterogeneity could happen by random chance. The key threshold here is typically 0.05. 
For example, if you get a p-value of 0.15, it suggests that the remaining variability between data points is likely just due to random differences (no significant heterogeneity).

Ideally, we hope that the Multivariate Cochran Q-test for residual heterogeneity to
not be significant (i.e. p >0.05), as this would indicate that high-heat days 
show consistent `r event` patterns (i.e., little variation from one high-heat day to the next)

A low, significant p-value indicates that there is more happening in the system than our model can capture. 

3) The AIC value (Akaike information criterion). 

We look to minimise AIC when all parts of the model are the same, except for the metapreds (listed in dlnm_setup_ns). Choose the model with the lower AIC. 

```{r summary, include = TRUE}

print(mvsummary)

``` 

```{r f-wald conditional clause, echo=FALSE, results="asis"}
if (length(metapreds) != 0) {
  cat("## Check of the f wald statistics for the meta predictors.\n\n")
  cat("We then want to see how good the metapredictors are that we chose for the model. Ideally, 
we are looking for the fwald to be significant.\n\n")
  cat("## Fwald for rangeweather\n\n")
  cat(paste0(rangeweather_fwald, "\n\n"))
  cat("## Fwald for maxweather\n\n")
  cat(paste0(maxweather_fwald, "\n\n"))
} else {
  cat("## There are no meta predictors in this model.\n\n")
}
```

## Check the residuals 

Ideally the box plots across the different y groupings should be in a horizontal line and the boxes should be approximately the same size. 

```{r box plot of residuals, echo=FALSE}

boxplot(mv$residuals)

```

## Check the residual autocorrelation

This matrix plot shows the patterns of correlation across the basis terms used in the model, assessing whether there are any significant dependencies remaining after accounting for the distributed lag effects.

In the matrix:

- Diagonal cells represent the autocorrelation of each basis term with itself over time, while off-diagonal cells show the cross-correlation between different basis terms at various time lags.
- If the residuals are well-behaved, most points should fall within the confidence bounds (typically shown as horizontal bands around zero), indicating no significant autocorrelation.

In our instance, the presence of high autocorrelation at lag 0 is not concerning. In DLNMs, basis terms are often functions (e.g., splines) representing smooth effects across a range of lags and exposure values. Some level of lag-0 correlation is expected because each basis term is derived from the same underlying exposure variable, creating some overlap or dependence at the immediate time point (lag 0).

When looking at the residuals, make sure that there aren't any plots with lags greater than 0 that breach a confidence interval.

```{r residual autocorrelation, fig.width=10, fig.height=10, echo=FALSE}

acf(resid(mv))

```
Ljung-Box significance test results for autocorrelation of residuals:

If the p-value > 0.05, the residuals are likely white noise.

```{r Ljung-Box test, echo=FALSE}

for (i in 1:ncol(mv$residuals)) {
  residuals_vector <- as.vector(mv$residuals[, i])
  boxlag <- floor(sqrt(length(residuals_vector)))
  cat("Ljung-Box test for column", i, "\n")
  print(Box.test(residuals_vector, lag = boxlag, type = "Ljung-Box"))
}

```
Augmented Dickey-Fuller (ADF) test to check if the residuals are stationary:

If the p-value is high (>0.05), the residuals are non-stationary.

```{r ADF test, echo=FALSE}

adf.test(residuals_vector)

```

## Optimal Weather Range compared to percentage thresholds

Below is the predicted optimal weather range (OWR) for each `r geog_area` based on the
mixmeta model. This should be the ideal temperatures to minimise risk of `r event`.

`r kable(OWR_thresholds)`

This is compared to the output for given percentage thresholds (ie. <2.5, 2.5-10% 90-97.5% and >97.5).

`r kable(perc_thresholds)`

If the output from OWR is close to the percentage thresholds, we will trial running the predictions for attributable number using OWR. 

## Thresholds used for assessing the temperatures which impact `r event`

We have established some key thresholds of importance for assessing which 
temperatures impact the likelihood of `r event`. These are the:

* **Minimum Mortality Threshold (MMT)** - the temperature at which the relative risk is equal to 1. At this point, a patient is not more or less likely to die. 

* **Risk Increase Threshold** - the temperature where the relative risk of death begins to increase significantly (i.e. the point at which the lower confidence interval for the RR curve is greater than 1).

* **High risk threshold** - the temperature at which the relative risk of death is equal to 1.1. This is the threshold used by England as the alert level.

* **Heatwave Day** - this is a single day that reaches 25 degrees celsius (the
established temperature for which 3 consecutive days would trigger a heatwave)

``` {r Thresholds, echo = FALSE}

thresholds <- c("Optimal Temperature Range", "Risk Increase", "High Risk", "Heatwave Day")

# Ensure 'temps' has exactly 4 elements
temps <- c(
  paste(owr_low, "-", owr_high),  # One string representing the range
  as.character(risk_increase),
  as.character(high_risk),
  25# 25 is included as high_risk already
)

# Create the dataframe
threshold_df <- data.frame(Threshold = thresholds, Temperature = temps, stringsAsFactors = FALSE)

kable(threshold_df)

```

## Scotland RR Plot

The below plot is calculated from the meta model across the regional models. It shows how the relative risk of `r event` changes across the summer temperatures. An RR of greater than 1 means that a person is more at risk from `r exposure`. I.e. at RR 1.1, a person is 10% more likely to be impacted. The vertical lines represent the thresholds detailed above.

```{r ScotRRPlot, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Relative risk across temperature in Scotland', fig.pos='H'}

Scotland_RR_Plot

```

## RR plots

```{r regionsRRPlot, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Relative risk across temperature in Scotland',  out.height = "700px", out.width='1000px', fig.pos='H'}

do.call(gridExtra::grid.arrange, plot_list)

```

## Plot of Scotland's weather

For reference when interpreting the attributable number note where the average maximum summer temperature lies.

```{r weather plot, echo=FALSE, warning=FALSE}
scot_weather_plot_mean

hotdays
```

## Attributable number `r event`

The following plots show the number of `r event` attributable to heat (high heat, risk_increase, high_risk and heatwave_day). These are calculated at regional level, and then the attributable number for Scotland is the sum across all regions. 

The main consideration from these plots is the size of the confidence intervals. Ideally we want these to be as narrow as possible, indicating that our estimates are more precise with 95% confidence. Also, if the lower CI remains above zero, we can say that there is attribution to `r exposure`.

# Scotland plots

## Attributable numbers

```{r scot_att_num_plots, echo=FALSE, warning=FALSE}

 gridExtra::grid.arrange(scot_att_plot_risk_increase,
                         scot_att_plot_high_risk,
                         scot_att_plot_heatwave_day, 
                         ncol = 1)

```

## Attributable rates

```{r scot_att_rate_plots, echo=FALSE}

gridExtra::grid.arrange(scot_att_rate_risk_increase,
                        scot_att_rate_high_risk,
                        scot_att_rate_heatwave_day, 
                        ncol = 1)

```


# Regional plots

## Regional Attributable Number plots

The following grid of plots shows the attributable `r event` at `r geog_area` level for summer `r event` over the risk_threshold. In other words, the charts 
show our modelled estimates for how many `r event` in each health board are attributable to heat each summer. The risk threshold is slightly different for each region - see RR plots for these thresholds. 

```{r reg_AN_risk_inc_plots, echo=FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = risk_increase_AN_plots, ncol = 3)
```

The below plots show the `r event` attributable to heat over the 1.1 RR threshold for the summer months each year, at health board level. 

```{r reg_AN_high_risk_plots, echo=FALSE, message = FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = high_risk_AN_plots, ncol = 3)
```

Finally, below plots show the `r event` attributable to heat on days at 25 degrees or higher during the summer months each year, at health board level.
 
```{r reg_AN_heatwave_day_plots, echo=FALSE, message = FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = heatwaveday_AN_plots, ncol = 3)
```

# Regional Attributable Rate Plots

## Risk increase

```{r reg_AR_risk_inc_plots, echo=FALSE, message = FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = risk_increase_AR_plots, ncol = 3)
```

## High risk

```{r reg_AR_high_risk_plots, echo=FALSE, message = FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = high_risk_AR_plots, ncol = 3)
```

## Heatwave day

```{r reg_AR_heatwave_day_plots, echo=FALSE, message = FALSE, warning=FALSE}
  gridExtra::grid.arrange(grobs = heatwaveday_AR_plots, ncol = 3)
```

