---
title: "Heat-related mortality/morbidity model using knots"
author: "Bella Tortora Brayda / Sarah Reed"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(knitr)
source(here("dlnm/dlnm_setup.R"))
source(here("dlnm/dlnm_code_knots.R"))
knitr::opts_chunk$set(echo = TRUE)
```

## DLNM Model Evaluation

### Context
 
A Distributed Lag Non-Linear Model (DLNM) is a prominent method used in population health to determine the effect of an exposure on a given outcome, allowing for a delay between exposure to an event and the resulting impact. 

In our context, we are evaluating how `r exposure` impacts `r event`.

For us to be able to accurately calculate the relative risk of `r event` from `r exposure`, and to assess the attributable `r event` for a given timeframe, we need a model that is robust. 

`r indicator_context`

This R markdown compiles the outputs created in the dlnm_code, and is designed to facilitate evaluating which, if any, of the model variations, produces robust estimates for attributable `r event` to `r exposure`.

## Parameters and variables of interest

This model variation evaluates the following independent columns: 

* `r independent_col1`
* `r independent_col2`
* `r independent_col3`
* `r independent_col4`

The parameters are as follows:

* Event = `r event`
* Exposure function (see dlnm::crossbasis) = `r varfun`
* Months included = `r month_choice`
* Percentile position of knots in exposure function (see dlnm::crossbasis) = `r varper` 
* Lag length in days = `r lag` 
* No. of knots in lag function = `r lagnk`
* Degrees of freedom for seasonality = `r dfseas` 
* Percentile used for centring predictions = `r percentile` 
* Minimum mortality/morbidity temperature = `r MMT`C

## Outputs of interest:

The main things to look for in the below output are:

1) Significance across the time periods (y1-y5)

This is indicated by small p-values (Pr(>|z|)) and for quick reference R signals these with stars. These indicate which variables significantly impact
mortality in our system.

2) Multivariate Cochran Q-test. 

This tells you the probability that the observed level of heterogeneity could happen by random chance. The key threshold here is typically 0.05. 
For example, if you get a p-value of 0.15, it suggests that the remaining variability between data points is likely just due to random differences (no significant heterogeneity).

Ideally, we hope that the Multivariate Cochran Q-test for residual heterogeneity to
not be significant (i.e. p >0.05), as this would indicate that high-heat days 
show consistent mortality patterns (i.e., little variation from one high-heat day to the next), 
even with the added weight, the p-value could increase, signaling that any remaining
heterogeneity is likely just due to random variation. In this case, you'd conclude
that there's no significant residual heterogeneity.

A low, significant p-value indicates that there is more happening in the system than our model can capture. 

3) The AIC value (Akaike information criterion). 

We look to minimise AIC when all parts of the model are the same, except for the metapreds (listed in dlnm_setup_ns). Choose the model with the lower AIC. 

```{r summary, include = TRUE}

print(mvsummary)

``` 

```{r f-wald conditional clause, echo=FALSE, results="asis"}
if (length(metapreds) != 0) {
  cat("## Check of the f wald statistics for the meta predictors.\n\n")
  cat("We then want to see how good the metapredictors are that we chose for the model. Ideally, 
we are looking for the fwald to be significant.\n\n")
  cat("## Fwald for rangeweather\n\n")
  cat(paste0(rangeweather_fwald, "\n\n"))
  cat("## Fwald for maxweather\n\n")
  cat(paste0(maxweather_fwald, "\n\n"))
} else {
  cat("## There are no meta predictors in this model.\n\n")
}
```

## Check the residuals 

Ideally the box plots across the different y groupings should be in a horizontal line and the boxes should be about the same size. 

```{r box plot of residuals, echo=FALSE}

boxplot(mv$residuals)

```

## Check the residual autocorrelation

This matrix plot shows the patterns of correlation across the basis terms used in the model, assessing whether there are any significant dependencies remaining after accounting for the distributed lag effects.

In the matrix:

- Diagonal cells represent the autocorrelation of each basis term with itself over time, while off-diagonal cells show the cross-correlation between different basis terms at various time lags.
- If the residuals are well-behaved, most points should fall within the confidence bounds (typically shown as horizontal bands around zero), indicating no significant autocorrelation.

In our instance, the presence of high autocorrelation at lag 0 is not concerning. In DLNMs, basis terms are often functions (e.g., splines) representing smooth effects across a range of lags and exposure values. Some level of lag-0 correlation is expected because each basis term is derived from the same underlying exposure variable, creating some overlap or dependence at the immediate time point (lag 0).

When looking at the residuals, make sure that there aren't any plots with lags greater than 0 that breach a confidence interval.

```{r residual autocorrelation, fig.width=10, fig.height=10, echo=FALSE}

acf(resid(mv))

```
Ljung-Box significance test results for autocorrelation of residuals:

If the p-value > 0.05, the residuals are likely white noise.

```{r Ljung-Box test, echo=FALSE}

for (i in 1:ncol(mv$residuals)) {
  residuals_vector <- as.vector(mv$residuals[, i])
  boxlag <- floor(sqrt(length(residuals_vector)))
  cat("Ljung-Box test for column", i, "\n")
  print(Box.test(residuals_vector, lag = boxlag, type = "Ljung-Box"))
}

```
Augmented Dickey-Fuller (ADF) test to check if the residuals are stationary:

If the p-value is high (>0.05), the residuals are non-stationary.

```{r ADF test, echo=FALSE}

adf.test(residuals_vector)

```

## Optimal Weather Range compared to percentage thresholds

Below is the predicted optimal weather range (OWR) for each healthboard based on the 
mv model. This should be the ideal temperatures to minimise risk of `r event`.

`r kable(OWR_thresholds)`

This is compared to the output for given percentage thresholds (ie. <2.5, 2.5-10% 90-97.5% and >97.5).

`r kable(perc_thresholds)`

If the output from OWR is close to the percentage thresholds, we will trial running the predictions for attributable number using OWR. 

Scotland's point of minimum mortality. Unless OWR aligns with percentage thresholds,
this will be at 75%. This is the point at which the RR is centred.

`r print(minperccountry)`

## Scotland RR Plot

The below plot is calculated from the meta model across the regional models. With this plot, we are looking that the confidence interval at the upper tail does not cross the RR line at 1. This indicates that with 95% confidence we can say that the RR of `r event` increases at a given threshold. The dashed lines indicate the 2.5th and 97.5th percentile of temperatures.

```{r ScotRRPlot, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Relative risk across temperature in Scotland', out.height = "460px", out.width='600px',fig.pos='H'}
knitr::include_graphics(paste0(climate_data_folder, indicator_output_path,
             scotRRplot,".pdf"))
```


## Plot of Scotland's weather

For reference when interpreting the attributable number note where the average maximum summer temperature lies.

```{r weather plot, echo=FALSE}

scot_weather_plot

```

## Attributable number and rate of deaths

The following plots show the number of deaths attributable to heat (high heat, risk_increase, high_risk and heatwave_day), followed by the attributable rate for high_heat. These are calculated at regional level, and then the attributable number/rate for Scotland is the sum across all regions. 

The general pattern should be the same, just the scale will be different. The main consideration from these plots is the size of the confidence intervals. Ideally we want these to be as narrow as possible, indicating that our estimates are more precise with 95% confidence. Also, if the lower CI remains above zero, we can say that there is attribution to `r exposure`.

```{r scot_att_plots, echo=FALSE}

 gridExtra::grid.arrange(scot_att_plot_risk_increase,
                         scot_att_plot_high_risk,
                         scot_att_plot_heatwave_day, 
                         ncol = 1)

```


# Regional plots

## RR plots

```{r regionsRRPlot, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Relative risk across temperature in Scotland', out.height = "460px", out.width='600px',fig.pos='H'}
knitr::include_graphics(paste0(climate_data_folder, indicator_output_path,
             regionsRRplot,".pdf"))

```

## Regional Attribution plots

The following grid of plots shows the attributable mortality at healthboard level for summer deaths over the risk_threshold. In other words, the charts 
show our modelled estimates for how many deaths in each health board are attributable to heat each summer. The risk threshold is slightly different for each region - see RR plots for these thresholds. 

```{r reg_att_plots_ri, echo=FALSE, message=FALSE}
risk_increase_plots <- map(region_names, plot_attribution_risk_inc) %>%
  gridExtra::grid.arrange(grobs = ., ncol = 3)

```

And finally, the below plots show the deaths attributable to heat on days 25 degrees or higher for the summer months each year, at health board level. 

```{r reg_att_plots_hw, echo=FALSE, message=FALSE}
heatwaveday_plots <- map(region_names, plot_attribution_heatwave_day) %>%
  gridExtra::grid.arrange(grobs = ., ncol = 3)

```
